---
title: "Statistical Tests"
output: html_document
---


```{r loading packages}
library(ggplot2)
library(ggthemes)
library(ggmap)
library(gridExtra)
library(scales)
library(car)
library(DescTools)
library(spdep)
library(readr)
library(dplyr)
```


```{r general data}
historical_data <- read_csv('data/processed/historical_data.csv')
future_data     <- read_csv('data/processed/future_data.csv')
snow_data       <- read_csv('data/processed/snow_data.csv') 

snow_data$month <- tolower(month.abb[as.numeric(snow_data$Month)])
snow_data$year  <- snow_data$Year
```


### Summerkill Statistics

```{r summer data}
fig1ab_data <- historical_data %>% 
  filter(month %in% c('jun', 'jul', 'aug', 'sep'))%>%
  mutate(cause.category.4 = replace(as.character(cause.category.4), which(is.na(cause.category.4)), 'Summer Non-event'))%>%
  mutate(infectious = ifelse(cause.category.4 == 'infectious agent', 1, 0))%>% 
  mutate(summerkill = ifelse(cause.category.4 == 'summerkill', 1, 0))
```

We use a Dunnett's test to compare the means of various kill categories against the mean of non kill lakes

```{r dunnet kill category ~ temperature}
DunnettTest(mean_surf~ factor(cause.category.4), data = fig1ab_data, control = 'Summer Non-event')
```

```{r normal distributionplot temp for non-events}
control_data <- fig1ab_data %>% 
  filter(cause.category.4 == 'Summer Non-event')
qqnorm(control_data$mean_surf)
```

```{r normal distribution plot temp for summerkill}
summerkill_data = fig1ab_data %>% filter(cause.category.4 == 'summerkill')
qqnorm(summerkill_data$mean_surf)
```

```{r normal distribution plot temp for infectious}
inf_data <- fig1ab_data %>%
  filter(cause.category.4 == 'infectious agent')
qqnorm(inf_data$mean_surf)
```

We use a Welch's t-test as opposed to a student's t-test to control for the uneven variance between sample groups when comparing event to non-event lakes directly 

```{r ttest summerkill~zscore}
ttest_1a  <- t.test(mean_surf_z ~ summerkill, data = fig1ab_data)
ttest_1a
```

```{r normal distribution plot zscore for summerkill}
summerkill_data <- fig1ab_data %>% filter(cause.category.4 == 'summerkill')
control_data <- fig1ab_data %>% filter(cause.category.4 == 'Summer Non-event')
qqplot(x = control_data$mean_surf_z, y = summerkill_data$mean_surf_z)
rm(summerkill_data)
```


```{r ttest infectious kill~zscore}
ttest_1b <- t.test(mean_surf_z ~ infectious, data = fig1ab_data)
ttest_1b
```

```{r normal distribution plot zscore for infectious vs non-events}
inf_data <- fig1ab_data %>% filter(cause.category.4 == 'infectious agent')
control_data <- fig1ab_data %>% filter(cause.category.4 == 'Summer Non-event')
qqplot(x = control_data$mean_surf_z, y = inf_data$mean_surf_z)
rm(inf_data)
rm(control_data)
```


### Winterkill Statistics

We run an ANOVA to compare the difference in mean temperature for winterkill vs non-event lakes

```{r anova winterkill~temperature}
fig1bd_data <- historical_data %>%
  filter(month %in% c('dec', 'jan', 'feb')) %>%
  mutate(cause.category.4 = replace(as.character(cause.category.4), which(is.na(cause.category.4)), 'Winter Non-event')) %>% 
  mutate(winterkill = ifelse(cause.category.4 == 'winterkill', 1, 0))

anova_data <- fig1bd_data  %>% 
  filter(cause.category.4 == 'Winter Non-event'| cause.category.4 == 'winterkill')
anova <- aov(mean_surf ~ factor(cause.category.4), anova_data)
summary(anova)
#tuk <- TukeyHSD(anova)
#tuk
```

```{r normal distributionplot temp for winter non-events}
control_data <- fig1bd_data %>%
  filter(cause.category.4 == 'Winter Non-event')
qqnorm(control_data$mean_surf)
```

```{r normal distribution plot temp for winterkill}
winterkill_data <- fig1bd_data %>% 
  filter(cause.category.4 == 'winterkill')
qqnorm(winterkill_data$mean_surf)
```

```{r normal distribution plot temp for winter vs non event}
qqplot(control_data$mean_surf, winterkill_data$mean_surf)
```

We use a Welch's t-test rather an a student due to unequal sizes between the two populations


```{r ttest winterkill~zscore}
ttest_1d <- t.test(mean_surf_z ~ winterkill, data = fig1bd_data)
ttest_1d
```

```{r normal distribution plot zscore for winter vs non event}
qqplot(control_data$mean_surf_z, winterkill_data$mean_surf_z)
rm(control_data)
rm(winterkill_data)
```


```{r anova winterkill~snow}
fig_SI3_data <- historical_data %>%   
  filter(month %in% c('dec', 'jan', 'feb'))  %>%
  mutate(lat_round = round(lat, 1), long_round = round(lon, 1)) %>%
  inner_join(snow_data, by = c('year', 'month', 'lat_round', 'long_round')) %>%
  mutate(winterkill = ifelse(!is.na(cause.category.4) & cause.category.4 == 'winterkill', 1, 0))

anova_snow <- aov(Snow ~ factor(winterkill), fig_SI3_data)
summary(anova_snow)
#tuk <- TukeyHSD(anova_snow)
#tuk
```

```{r normal distribution plot snow for winterkill vs non event}
control_data <- fig_SI3_data %>%
  filter(winterkill == 0)
winterkill_data <- fig_SI3_data %>% 
  filter(winterkill == 1)

qqplot(control_data$Snow, winterkill_data$Snow)

rm(control_data)
rm(winterkill_data)

```
We employ a one sample t-test to see if the mean zscore is different from zero

```{r ttest winterkill~zscore all seasons}
ttest_data_z <- filter(historical_data, cause.category.4 == 'winterkill')
t.test(ttest_data_z$mean_surf_z, mu = 0)
```



```{r anova winterkill~ice duration}
fig_SI4_data <- historical_data  %>%
  mutate(winterkill = ifelse(!is.na(cause.category.4) & 
                               cause.category.4 == 'winterkill', 1, 0)) %>%
  group_by(year, wbic) %>%
  summarise(ice_duration = mean(ice_duration), winterkill = max(winterkill))

anova <- aov(ice_duration ~ winterkill, fig_SI4_data)
summary(anova)
#tuk <- TukeyHSD(anova)
#tuk
```


```{r normal distribution plot ice_duration for winterkill vs non event}
control_data <- fig_SI4_data %>% filter(winterkill == 0)
winterkill_data <- fig_SI4_data %>% filter(winterkill == 1)
qqplot(control_data$ice_duration, winterkill_data$ice_duration)
rm(control_data)
rm(winterkill_data)
```


### Moran Tests

```{r}
fig3a_data <-historical_data %>%
  mutate(summerkill_binary = ifelse(is.na(summerkill) | summerkill == 'neg', 0 , 1)) %>%
  group_by(wbic) %>%
  summarise(MME = max(summerkill_binary), V1 = mean(lon), V2 = mean(lat))

fig3b_data <- future_data %>%
  dplyr::select(wbic, lat, lon, year) %>%
  bind_cols(predictions_1) %>%
  mutate(event = pos) %>%
  mutate(no_event_prob = 1 - event) %>%
  filter(year > 2030 & year < 2070) %>%
  group_by(wbic) %>%
  summarise(prob = (1 - prod(no_event_prob)), V1 = mean(lon), V2 = mean(lat)) 

fig3c_data <- future_data %>%
  dplyr::select(wbic, lat, lon, year) %>%
  bind_cols(predictions_1)%>%
  mutate(event = pos) %>%
  mutate(no_event_prob = 1 - event) %>%
  filter(year > 2070) %>%
  group_by(wbic) %>%
  summarise(prob = (1 - prod(no_event_prob)), V1 = mean(lon), V2 = mean(lat)) 

coordinates(fig3a_data) <- c('V1', 'V2')
coordinates(fig3b_data) <- c('V1', 'V2')
coordinates(fig3c_data) <- c('V1', 'V2')

nb_p1 <- knn2nb(knearneigh(coordinates(coordinates(fig3a_data)),longlat = TRUE, k = 2))
nb_p2 <- knn2nb(knearneigh(coordinates(coordinates(fig3b_data)),longlat = TRUE, k = 2))
nb_p3 <- knn2nb(knearneigh(coordinates(coordinates(fig3c_data)),longlat = TRUE, k = 2))

weights_p1 <- nb2listw(nb_p1)
weights_p2 <- nb2listw(nb_p2)
weights_p3 <- nb2listw(nb_p3)

moran_MME   <- moran.test(fig3a_data$MME, weights_p1)
moran_Prob2 <- moran.test(fig3b_data$prob, weights_p2)
moran_Prob3 <- moran.test(fig3c_data$prob, weights_p3)

moran_MME
moran_Prob2
moran_Prob3
```